INGRID: Creating Languages in MPS from ANTLR Grammars

### 2017-06-05 chapters/background.tex
todo

@@

The building blocks of MPS models are called nodes.
Code of any program in MPS is built from nodes, which represent instances of concepts defined in one of the languages that the program declares to be using.
In the context of MPS, a concept is, in fact, a building block of a language definition.
We use the terms \emph{MPS concept} and \emph{AST node} when needed to avoid confusion.

@@

>>>>

\todo{For each MPS term (Structure, Editor, TextGen, concept, and so on) say which generally/widely known terms (from related academic work) are similar/analogous. Relate the MPS-specific terms to the large body of academic work.}

@@

The building blocks of MPS models are called nodes.
Code of any program in MPS is built from nodes, which represent instances of concepts defined in one of the languages that the program declares to be using.
\todo{zvyraznit slovo "concept"} In the context of MPS, a concept is, in fact, a building block of a language definition.
We use the terms \emph{MPS concept} and \emph{AST node} when needed to avoid confusion.

@@

\todo{
S. Erdweg, T. Rendel, C. K\"{a}stner, and K. Ostermann. Layout-Sensitive Generalized Parsing. In Proceedings of SLE 2012, LNCS 7745.
	Source code layout specified declaratively through constraints on shape and relative positions of syntax trees (AST).
	Layout constraints are represented/encoded using annotations on production rules of context-free grammars.
	Useful for layout-sensitive programming language (e.g., Python).
	Their proposed approach: grammar rule applies only if the input satisfied the constraint (on shape and positioning of source code fragments).
		Parsing algorithm that accepts only inputs with valid layout (that satisfies the constraints attached to grammar rules).
	Constraints support (can express): indentation, number of line breaks, alignment.
	Hand-coded approach used in Python (how the annotated grammar is processed) --- mentioned for comparison.
		Special tokens that indicate layout aspects (indentation, alignment) are added to stream of tokens by preprocessor.
		Layout handled by algorithm that works with syntax trees (ASTs) and token streams.
	The main difference from our approach (INGRID): layout inferred automatically (via automated learning) from several input files.
		No layout annotations in the grammar are used.
		Currently, simple heuristics are used to create editor and textgen.
}

### 2017-06-05 chapters/conclusion.tex
small changes
todo

Our contribution is the INGRID method for constructing language definitions in MPS from their grammars in the ANTLR notation.
The method is only partially automated, but nevertheless it greatly reduces the amount of manual work that developers of MPS languages have to perform.
Although we discussed all the challenges, details of our solution and general observations just in the context of ANTLR grammars and JetBrains MPS, we believe they are more general --- and in particular relevant to everyone tackling a similar problem in the field of DSLs and language workbenches.

In the future, we would like to address the remaining open problems (there are many of them) and increase the level of automation.
Our primary research goal is to investigate the usage of machine learning for the purpose of generating editors that support nice and readable code layout.
The information about code layout, once it is available, will be then leveraged and used also to improve TextGen.
Another subject for future work is generating parsers that would allow users to import existing source code into MPS.

>>>>

Our contribution is the \textsc{Ingrid} method for constructing language definitions in MPS from their grammars in the ANTLR notation.
The method is only partially automated, but nevertheless it greatly reduces the amount of manual work that developers of MPS languages have to perform.
Although we discussed all the challenges, details of our solution and general observations just in the context of ANTLR grammars and JetBrains MPS, we believe they are more general --- and in particular relevant to everyone tackling a similar problem in the field of DSLs and language workbenches.

In the future, we would like to address the remaining open problems (there are many of them) and increase the level of automation.
Our primary research goal is to investigate the usage of machine learning for the purpose of generating editors that support nice and readable code layout.
The information about code layout, once it is available, will be then leveraged and used also to improve TextGen.
Another subject for future work is generating parsers that would allow users to import existing source code into MPS \todo{Vysvetlit podrobneji ze chceme automaticky generovat parsery textovych zdrojaku ktere budou vytvaret interni reprezentace (ASTs) postavene na zaklade MPS concepts pro dany jazyk}.

### 2017-06-05 chapters/introduction.tex
todo

In this paper we present \textsc{Ingrid}, a method and a tool for semi-automated construction of the definition of a programming language in MPS that uses an ANTLR grammar of the language as a description of its syntax.
We discuss the main challenges that we encountered during the development of \textsc{Ingrid} and our solution to them.
Although we explain everything on the examples of MPS languages and ANTLR grammars, we emphasize that most of the challenges are general. 
They apply to the relation between the two approaches to the definition of a programming language --- (i) one that uses grammars in an EBNF-like notation, with rules and tokens written in plain text, and the corresponding parser generators, and (ii) the structured object-oriented approach used in MPS and other language workbenches.
Specifically, the description of a language in the form of a grammar does not hold any information about the code layout, and the grammar typically contains many rules that do not directly correspond to AST nodes and programming language constructs.
We provide more details in Section~\ref{sect:LANGDEF}.

>>>>

In this paper we present \textsc{Ingrid}, a method and a tool for semi-automated construction of the definition of a programming language in MPS that uses an ANTLR grammar of the language as a description of its syntax.
We discuss the main challenges that we encountered during the development of \textsc{Ingrid} and our solution to them.
\todo{Zmenit: we explain our tool using examples from MPS and ANTLR} Although we explain everything on the examples of MPS languages and ANTLR grammars, we emphasize that most of the challenges are general. 
They apply to the relation between the two approaches to the definition of a programming language --- (i) one that uses grammars in an EBNF-like notation, with rules and tokens written in plain text, and the corresponding parser generators, and (ii) the structured object-oriented approach used in MPS and other language workbenches.
Specifically, the description of a language in the form of a grammar does not hold any information about the code layout, and the grammar typically contains many rules that do not directly correspond to AST nodes and programming language constructs.
We provide more details in Section~\ref{sect:LANGDEF}.

### 2017-06-05 chapters/language_definition.tex
todo

The proposed \textsc{Ingrid} method accepts grammars in the ANTLR v4 notation~\cite{ref:ANTLRBOOK} as input.
All widely used programming languages have their syntax defined using this notation~\footnote{https://github.com/antlr/grammars-v4}.
Unlike some of the related projects, we did not extend the notation with any custom features, and we also did not create an MPS language for the ANTLR notation.

@@

Every tree that represents a lexer rule is flattened into the equivalent regular expression by the recursive algorithm in Figure~\ref{fig:ALGFLATTEN}.
A sequence is created from the elements of each alternative, and then all those sequences are joined by the alternation operator \code{$|$} to form a regular expression.

@@

When designing the process for translating AST nodes (i.e., grammar rules) into the MPS language structure, we faced several challenges.
The main challenge related to the structure, as we already mentioned in the introduction, is that a grammar typically contains rules that do not directly correspond to programming language constructs.
Such rules exist at the intermediate layers of a syntax hierarchy.
Their main purpose is to enable easier understanding and maintenance of the grammar by humans.
However, presence of the intermediate layers significantly complicates usage of the given language in MPS, and the layers are actually not necessary for construction of an MPS language.
We show examples illustrating this problem, which we call a \emph{layer problem}, later in the subsection where we describe our solution.

@@

For the input that consists of a particular AST node $N$ and the grammar rule $R$ that expands $N$, the procedure implementing the shortcut approach systematically traverses the parser tree (AST) built in the earlier phases in order to identify each AST node that may appear at the end of some derivation chain starting by the rule $R$ from the node $N$.
Such nodes cannot expand further based on any rule in the grammar, and for that reason we call them \emph{end nodes}.

@@

We experimented with several heuristics to derive a useful code layout, but all of them produced rather suboptimal results.
However, we observed that the most tedious and error-prone step in the manual definition of the Editor aspect is the creation of cells for all literals (keywords), properties, children, and other fields of a given concept that should appear in its visual representation.
This step can be very easily automated.

@@

We have found that the combination of two steps, (1) automated placing of cells into a single row and (2) subsequent manual adjustment of the layout, is a very fast and efficient way of creating a nice code layout.
Nevertheless, we plan to work on a more automated approach to the definition of Editor aspects in the future, probably using some techniques of machine learning.
The key idea would be to derive the code layout from a set of valid input source files.
We believe that the user would have to adjust the editor manually a little bit even in this case, because the results of a machine learning-based approach would not be perfect.

@@

Our \textsc{Ingrid} method targets mostly text-based languages, for the concepts of which the visual representation in a projectional editor must be almost equivalent to their plain-text representation.
An obvious choice would therefore be to use the same approach for Editor and TextGen.
Nevertheless, since the Editor aspect has to be adjusted manually, we decided to use a different approach for TextGen.
We designed a procedure based on a simple fully automated heuristic that provides surprisingly good results.

@@

Figure~\ref{fig:TEXTGENFINAL} contains the complete, automatically generated, TextGen aspect for the \mpsconcept{Element{\_}1} concept, which represents the full SimpleXML element.
Such code may be adjusted manually very easily in order to produce nicely indented XML documents.
We only need to wrap the \mpsconcept{Content{\_}2} child node with indentation and change the sequence separator to a new line character.
A fragment of the resulting adjusted aspect (code) is in Figure~\ref{fig:TEXTGENADJUSTED}.

>>>>

The proposed \textsc{Ingrid} method accepts grammars in the ANTLR v4 notation~\cite{ref:ANTLRBOOK} as input.
\todo{Zmenit nasledujici vetu: There exist ANTLR v4 grammars for all widely used programming languages ...}
All widely used programming languages have their syntax defined using this notation~\footnote{https://github.com/antlr/grammars-v4}.
Unlike some of the related projects, we did not extend the notation with any custom features, and we also did not create an MPS language for the ANTLR notation.

@@

Every tree that represents a lexer rule is flattened into the equivalent regular expression by the recursive algorithm in Figure~\ref{fig:ALGFLATTEN}.
A sequence is created from the elements of each alternative, and then all those sequences are joined by the alternation operator \code{$|$} to form a regular expression.
\todo{Algorithm seems to fail if the grammar has right-recursion (according to one reviewer) --- check that and describe how we handle that case.}

@@

When designing the process for translating AST nodes (i.e., grammar rules) into the MPS language structure, we faced several challenges.
The main challenge related to the structure, as we already mentioned in the introduction, is that a grammar typically contains rules that do not directly correspond to programming language constructs.
Such rules exist at the intermediate layers of a syntax hierarchy.
Their main purpose is to enable easier understanding and maintenance of the grammar by humans.
However, presence of the intermediate layers significantly complicates usage of the given language in MPS, and the layers are actually not necessary for construction of an MPS language.
We show examples illustrating this problem, which we call a \emph{layer problem}, later in the subsection where we describe our solution \todo{Pridat odkaz na konkretni section a opravit gramatickou chybu}.

@@

For the input that consists of a particular AST node $N$ and the grammar rule $R$ that expands $N$, the procedure implementing the shortcut approach systematically traverses the parser tree (AST) built in the earlier phases in order to identify each AST node that may appear at the end of some derivation chain starting by the rule $R$ from the node $N$.
Such nodes cannot expand further based on any rule in the grammar, and for that reason we call them \emph{end nodes}. \todo{Srovnat pojem "end nodes" vzhledem ke terminal symbols from grammars (jako analogie). Mozna existuje dalsi vhodny termin.}

@@

We experimented with several heuristics to derive a useful code layout, but all of them produced rather suboptimal results \todo{Give examples or quantify this result}.
However, we observed that the most tedious and error-prone step in the manual definition of the Editor aspect is the creation of cells for all literals (keywords), properties, children, and other fields of a given concept that should appear in its visual representation.
This step can be very easily automated.

@@

We have found that the combination of two steps, (1) automated placing of cells into a single row and (2) subsequent manual adjustment of the layout, is a very fast and efficient way of creating a nice code layout.
Nevertheless, we plan to work on a more automated approach to the definition of Editor aspects in the future, using automated learning.
The key idea would be to derive the code layout from a set of valid input source files.
We believe that the user would have to adjust the editor manually a little bit even in this case, because the results of a machine learning-based approach would not be perfect.

@@

Our \textsc{Ingrid} method targets mostly text-based languages, for the concepts of which the visual representation in a projectional editor must be almost equivalent to their plain-text representation.
An obvious choice would therefore be to use the same approach for Editor and TextGen.
Nevertheless, since the Editor aspect has to be adjusted manually, we decided to use a different approach for TextGen.
We designed a procedure based on a simple fully automated heuristic that provides surprisingly good results \todo{Quantify this}.

@@

\todo{Figure 15: whitespace is just blank space instead of quoted string, mozna jsem neco smazal omylem (za "with ").}
Figure~\ref{fig:TEXTGENFINAL} contains the complete, automatically generated, TextGen aspect for the \mpsconcept{Element{\_}1} concept, which represents the full SimpleXML element.
Such code may be adjusted manually very easily in order to produce nicely indented XML documents.
We only need to wrap the \mpsconcept{Content{\_}2} child node with indentation and change the sequence separator to a new line character.
A fragment of the resulting adjusted aspect (code) is in Figure~\ref{fig:TEXTGENADJUSTED}.
